Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2019-04-12 WholePlotTrajectories.Rmd
relecture

An early conceptual basis of the linkage between biodiversity and disturbance is the Intermediate Disturbance Hypothesis (IDH).
The IDH assumes a relationship between community diversity and the intensity and frequency of disturbance events, stating that community diversity peaks at intermediate level of disturbance [@Connell1978].
The theory assumes that the environmental fluctuations following disturbance foster both competitively superior species and fast colonizers and thus prevent competitive exclusion [@Shea2004;@Pulsford2016].
In tropical forests, however, observations of the IDH often diverge from theoretical expectations [@Hugues2007;@Sheil2003;@Norden2017], and the validation of the theory remains controversial [@Hubbell2001;@Fox2013;@Sheil2013].

Moreover, we argue that there is a need to delineate the scope of the IDH in time, using ecosystem time-explicit trajectories, and highlight the moment when the conditions for the IDH emerge [@Sheil2003;@Shea2004]. Indeed, the link between biodiversity, IDH
and post-disturbance disturbance trajectories might be complicated by the huge diversity found in tropical
tree communities and the variety of ecological processes that may underlie the IDH
(colonization, facilitation, coexistence,
etc)[@Lindenmayer2012;@Garcia_florez2017].

Analysing community response to disturbance requires a set of metrics, to grasp all aspects of community changes [@Sheil2003;@Shea2004;@Mayfield2010].
The analysis should first consider community composition, which is crucial for conservation issues and which reveals the pool of species fostered (or hampered) by disturbance [@Lavorel2002;@Bellwood2006].
Then, diversity metrics encompassing both richness and evenness should also be considered to assess the changes in community abundance distribution. 
Besides, functional approaches have been shown to usefully complement pure taxonomic approaches because they shed light on the species biological attributes that directly link community diversity, composition and redundancy to ecosystem functioning [@Violle2007b;@Baraloto2012a]. 
In that respect, a vast literature allowed recognizing major traits that represent species ecological strategy and determine how they respond to changing conditions [@Diaz2005]. 
Specifically in tropical forests, the functional approach revealed the post-disturbance deterministic processes entailing a shift from a dominance of "conservative" slow-growing species dealing with scarce resources, to a dominance of "acquisitive" fast-growing species with rapid and efficient use of abundant resources [@Rees2001;@Reich2014;@Herault2011]. 
This shift is translated into the trajectories of average community value of key functional traits related to resource acquisition (leaf and stem traits) and life-history strategy (seed mass, maximum size) [@Wright2004;@TerSteege2006;@Westoby2006a;@Chave2009b].

@@

Community trajectories depended on the disturbance intensity, with a threshold of 20-25% of AGB loss for which the taxonomic richness was maximized and the taxonomic evenness remained resilient.
Below a disturbance intensity threshold (treatments 1 and 2), the taxonomic richness increased and the taxonomic eveness increased first before stabilizing or starting to decrease.
In those cases, the trees surviving after disturbance remained numerous enough to maintain the high taxonomic richness of the pre-disturbance community [@Bongers2009].
In parallel, the recruitment of pioneers, infrequent or absent before disturbance, increased the taxonomic richness all the more so that the disturbance was intense [@Martin2015;@Chaudhary2016].
As these pioneers became more dominant they balanced the usual hyper-dominance of tropical forests and increased the taxonomic evenness [@Baraloto2012a].
Contrastingly, above the intensity threshold (treatment 3), the taxonomic richness did not exceed the initial value and the taxonomic evenness continuously increased without sign of reversal.
Above the intensity threshold, the taxonomic richness of surviving trees was too low to be offset by the recruitment of pioneers. 
In the Guiana shield indeed, the pool of true pioneers specifically recruited after disturbance is restricted to a few common genera (e.g. *Cecropia* spp., *Vismia* spp.) [@Guitet2018].
At the same time, pioneers became persistently dominant and prevented a return towards the initial evenness and the recovery of hyper-dominant shade-tolerant species.

After disturbance of intermediate intensity, taxonomic evenness trajectories were hump-shaped, as already observed in the Guiana shield [@Baraloto2012a] and in Bornean tropical forests [@Cannon1998].
Trajectories peaked at an intermediate time after disturbance, around 15-20 years.
This time corresponded to a balance between first established pioneers and the more competitve late-successional species.

Regarding community functional trajectories in contrast, there was neither intermediate disturbance intensity nor intermediate time after disturbance, which dismissed the IDH.
Neither functional richness nor functional evenness displayed a humped-shaped pattern, but both rather increased with the recruitment of pioneers that were functionally highly different from the pre-disturbance community [@Denslow1980;@Molino2001].
Above the intensity threshold however, for the most intense disturbance, community functional richness and evenness started to decrease after 15 to 20 years. 
Right after disturbance, short-lived species benefited from the resources made available and prevented the establishment of other species. 
The decline of these short-lived pioneers decreased the functional richness and evenness, but we suggest that the establishment of long-lasting pioneers will follow and that the trajectories will catch up with those observed for intermediate disturbance [@Walker2009].

>>>>

An early conceptual basis of the linkage between biodiversity and disturbance is the Intermediate Disturbance Hypothesis (IDH).
The IDH assumes a relationship between community diversity and the intensity and frequency of disturbance events, stating that community diversity peaks at intermediate level of disturbance [@Connell1978].
The theory assumes that the environmental fluctuations following disturbance foster both competitively superior species and fast colonizers and thus prevent competitive exclusion [@Shea2004;@Pulsford2016].
In tropical forests, however, observations of the IDH often diverge from theoretical expectations [@Hugues2007;@Sheil2003;@Norden2017], and the validation of the theory remains controversial [@Hubbell2001;@Fox2013;@Sheil2013].
In tropical forests the processes underlying the IDH might indeed be complicated by the huge diversity found in tropical tree communities and the variety of ecological processes that may underlie the IDH
(colonization, facilitation, coexistence, etc) [@Lindenmayer2012;@Garcia_florez2017].
It is then crucial to test the IDH in this context and determine its scope in time, using ecosystem time-explicit trajectories [@Sheil2003;@Shea2004].

Analysing community response to disturbance requiers a set of metrics, to grasp all aspects of community changes [@Sheil2003;@Shea2004;@Mayfield2010].
The analysis should first consider community composition, which is crucial for conservation issues and which reveals the pool of species fostered (or hampered) by disturbance [@Lavorel2002;@Bellwood2006].
Then, diversity metrics encompassing both richness and evenness should also be considered to assess the changes in community abundance distribution. 
Besides, functional approaches have been shown to usefully complement pure taxonomic approaches because they shed light on the species biological attributes that directly link community diversity, composition and redundancy to ecosystem functioning [@Violle2007b;@Baraloto2012a]. 
In that respect, a vast literature allowed recognizing major traits that represent species ecological strategy and determine how they respond to changing conditions [@Diaz2005]. 
Specifically in tropical forests, the functional approach revealed the post-disturbance deterministic processes entailing a shift from a dominance of "conservative" slow-growing species dealing with scarce resources, to a dominance of "acquisitive" fast-growing species with rapid and efficient use of abundant resources [@Rees2001;@Reich2014;@Herault2011]. 
This shift is translated into the trajectories of average community value of key functional traits related to resource acquisition (leaf and stem traits) and life-history strategy (seed mass, maximum size) [@Wright2004;@TerSteege2006;@Westoby2006a;@Chave2009b].

@@

Trajectories of taxonomic richness and evenness were markedly different below and above an intensity threshold (\ref{fig:DivTaxo}).
Below the intensity threshold, for disturbances in the order of treatments 1 and 2, both taxonomic richness and evenness increased according to the disturbance intensity [@Martin2015;@Chaudhary2016].
This suggested that while trees surviving after disturbance remained numerous enough to maintain the richness of the pre-disturbance community [@Bongers2009], the recruitment of pioneers previously infrequent or absent allowed increasing the taxonomic richness [@Martin2015;@Chaudhary2016].
As these pioneers became more dominant, they balanced the usual hyper-dominance of tropical forests and increased the taxonomic evenness [@Baraloto2012a].
Above the intensity threshold, like for treatment 3, the taxonomic richness did not exceed the initial value.
The richness of surviving trees was then too low to be offset by the recruitment of pioneers. 
In the Guiana shield indeed, the pool of true pioneers specifically recruited after disturbance is restricted to a few common genera (e.g. *Cecropia* spp., *Vismia* spp.) [@Guitet2018].

Considering different moments of the post-disturbance trajectories, there was always a humped-shaped pattern linking taxonomic richness or evenness and the intensity of disturbance (\ref{fig:IDHplot}). 
Both taxonomic richness and evenness were maximized at an intermediate intensity, around 20-25% of AGB lost, and this along the 30 years following disturbance.

Regarding community functional trajectories, no marked differences were observed among the post-disturbance trajectories (\ref{fig:DivTaxo}).
Whenever the disturbance intensity, functional richness and evenness first increased with the recruitment of pioneers that were functionally highly different from the pre-disturbance community [@Denslow1980;@Molino2001].
Then, after 15 to 20 years community functional richness and evenness started to decrease, with the decline of the first established species.
Right after disturbance, short-lived pioneers benefited from the resources made available and prevented the establishment of other species. 
These pioneers then declined, which decreased the functional richness and evenness [@Walker2009].

Considering different times after disturbance, both functional richness and evenness kept increasing with the disturbance intensity and did not display humped-shaped trajectories (\ref{fig:IDHplot}).

### 2019-04-12 WholePlotTrajectories.tex
relecture

An early conceptual basis of the linkage between biodiversity and
disturbance is the Intermediate Disturbance Hypothesis (IDH). The IDH
assumes a relationship between community diversity and the intensity and
frequency of disturbance events, stating that community diversity peaks
at intermediate level of disturbance \citep{Connell1978}. The theory
assumes that the environmental fluctuations following disturbance foster
both competitively superior species and fast colonizers and thus prevent
competitive exclusion \citep{Shea2004, Pulsford2016}. In tropical
forests, however, observations of the IDH often diverge from theoretical
expectations \citep{Hugues2007, Sheil2003, Norden2017}, and the
validation of the theory remains controversial
\citep{Hubbell2001, Fox2013, Sheil2013}.

Moreover, we argue that there is a need to delineate the scope of the
IDH in time, using ecosystem time-explicit trajectories, and highlight
the moment when the conditions for the IDH emerge
\citep{Sheil2003, Shea2004}. Indeed, the link between biodiversity, IDH
and post-disturbance disturbance trajectories might be complicated by
the huge diversity found in tropical tree communities and the variety of
ecological processes that may underlie the IDH (colonization,
facilitation, coexistence,
etc)\citep{Lindenmayer2012, Garcia_florez2017}.

Analysing community response to disturbance requires a set of metrics,
to grasp all aspects of community changes
\citep{Sheil2003, Shea2004, Mayfield2010}. The analysis should first
consider community composition, which is crucial for conservation issues
and which reveals the pool of species fostered (or hampered) by
disturbance \citep{Lavorel2002, Bellwood2006}. Then, diversity metrics
encompassing both richness and evenness should also be considered to
assess the changes in community abundance distribution. Besides,
functional approaches have been shown to usefully complement pure
taxonomic approaches because they shed light on the species biological
attributes that directly link community diversity, composition and
redundancy to ecosystem functioning \citep{Violle2007b, Baraloto2012a}.
In that respect, a vast literature allowed recognizing major traits that
represent species ecological strategy and determine how they respond to
changing conditions \citep{Diaz2005}. Specifically in tropical forests,
the functional approach revealed the post-disturbance deterministic
processes entailing a shift from a dominance of ``conservative''
slow-growing species dealing with scarce resources, to a dominance of
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{Rees2001, Reich2014, Herault2011}. This shift
is translated into the trajectories of average community value of key
functional traits related to resource acquisition (leaf and stem traits)
and life-history strategy (seed mass, maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}.

@@

Community trajectories depended on the disturbance intensity, with a
threshold of 20-25\% of AGB loss for which the taxonomic richness was
maximized and the taxonomic evenness remained resilient. Below a
disturbance intensity threshold (treatments 1 and 2), the taxonomic
richness increased and the taxonomic eveness increased first before
stabilizing or starting to decrease. In those cases, the trees surviving
after disturbance remained numerous enough to maintain the high
taxonomic richness of the pre-disturbance community \citep{Bongers2009}.
In parallel, the recruitment of pioneers, infrequent or absent before
disturbance, increased the taxonomic richness all the more so that the
disturbance was intense \citep{Martin2015, Chaudhary2016}. As these
pioneers became more dominant they balanced the usual hyper-dominance of
tropical forests and increased the taxonomic evenness
\citep{Baraloto2012a}. Contrastingly, above the intensity threshold
(treatment 3), the taxonomic richness did not exceed the initial value
and the taxonomic evenness continuously increased without sign of
reversal. Above the intensity threshold, the taxonomic richness of
surviving trees was too low to be offset by the recruitment of pioneers.
In the Guiana shield indeed, the pool of true pioneers specifically
recruited after disturbance is restricted to a few common genera (e.g.
\emph{Cecropia} spp., \emph{Vismia} spp.) \citep{Guitet2018}. At the
same time, pioneers became persistently dominant and prevented a return
towards the initial evenness and the recovery of hyper-dominant
shade-tolerant species.

After disturbance of intermediate intensity, taxonomic evenness
trajectories were hump-shaped, as already observed in the Guiana shield
\citep{Baraloto2012a} and in Bornean tropical forests
\citep{Cannon1998}. Trajectories peaked at an intermediate time after
disturbance, around 15-20 years. This time corresponded to a balance
between first established pioneers and the more competitve
late-successional species.

Regarding community functional trajectories in contrast, there was
neither intermediate disturbance intensity nor intermediate time after
disturbance, which dismissed the IDH. Neither functional richness nor
functional evenness displayed a humped-shaped pattern, but both rather
increased with the recruitment of pioneers that were functionally highly
different from the pre-disturbance community
\citep{Denslow1980, Molino2001}. Above the intensity threshold however,
for the most intense disturbance, community functional richness and
evenness started to decrease after 15 to 20 years. Right after
disturbance, short-lived species benefited from the resources made
available and prevented the establishment of other species. The decline
of these short-lived pioneers decreased the functional richness and
evenness, but we suggest that the establishment of long-lasting pioneers
will follow and that the trajectories will catch up with those observed
for intermediate disturbance \citep{Walker2009}.

>>>>

An early conceptual basis of the linkage between biodiversity and
disturbance is the Intermediate Disturbance Hypothesis (IDH). The IDH
assumes a relationship between community diversity and the intensity and
frequency of disturbance events, stating that community diversity peaks
at intermediate level of disturbance \citep{Connell1978}. The theory
assumes that the environmental fluctuations following disturbance foster
both competitively superior species and fast colonizers and thus prevent
competitive exclusion \citep{Shea2004, Pulsford2016}. In tropical
forests, however, observations of the IDH often diverge from theoretical
expectations \citep{Hugues2007, Sheil2003, Norden2017}, and the
validation of the theory remains controversial
\citep{Hubbell2001, Fox2013, Sheil2013}. In tropical forests the
processes underlying the IDH might indeed be complicated by the huge
diversity found in tropical tree communities and the variety of
ecological processes that may underlie the IDH (colonization,
facilitation, coexistence, etc)
\citep{Lindenmayer2012, Garcia_florez2017}. It is then crucial to test
the IDH in this context and determine its scope in time, using ecosystem
time-explicit trajectories \citep{Sheil2003, Shea2004}.

Analysing community response to disturbance requiers a set of metrics,
to grasp all aspects of community changes
\citep{Sheil2003, Shea2004, Mayfield2010}. The analysis should first
consider community composition, which is crucial for conservation issues
and which reveals the pool of species fostered (or hampered) by
disturbance \citep{Lavorel2002, Bellwood2006}. Then, diversity metrics
encompassing both richness and evenness should also be considered to
assess the changes in community abundance distribution. Besides,
functional approaches have been shown to usefully complement pure
taxonomic approaches because they shed light on the species biological
attributes that directly link community diversity, composition and
redundancy to ecosystem functioning \citep{Violle2007b, Baraloto2012a}.
In that respect, a vast literature allowed recognizing major traits that
represent species ecological strategy and determine how they respond to
changing conditions \citep{Diaz2005}. Specifically in tropical forests,
the functional approach revealed the post-disturbance deterministic
processes entailing a shift from a dominance of ``conservative''
slow-growing species dealing with scarce resources, to a dominance of
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{Rees2001, Reich2014, Herault2011}. This shift
is translated into the trajectories of average community value of key
functional traits related to resource acquisition (leaf and stem traits)
and life-history strategy (seed mass, maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}.

@@

Trajectories of taxonomic richness and evenness were markedly different
below and above an intensity threshold (\ref{fig:DivTaxo}). Below the
intensity threshold, for disturbances in the order of treatments 1 and
2, both taxonomic richness and evenness increased according to the
disturbance intensity \citep{Martin2015, Chaudhary2016}. This suggested
that while trees surviving after disturbance remained numerous enough to
maintain the richness of the pre-disturbance community
\citep{Bongers2009}, the recruitment of pioneers previously infrequent
or absent allowed increasing the taxonomic richness
\citep{Martin2015, Chaudhary2016}. As these pioneers became more
dominant, they balanced the usual hyper-dominance of tropical forests
and increased the taxonomic evenness \citep{Baraloto2012a}. Above the
intensity threshold, like for treatment 3, the taxonomic richness did
not exceed the initial value. The richness of surviving trees was then
too low to be offset by the recruitment of pioneers. In the Guiana
shield indeed, the pool of true pioneers specifically recruited after
disturbance is restricted to a few common genera (e.g. \emph{Cecropia}
spp., \emph{Vismia} spp.) \citep{Guitet2018}.

Considering different moments of the post-disturbance trajectories,
there was always a humped-shaped pattern linking taxonomic richness or
evenness and the intensity of disturbance (\ref{fig:IDHplot}). Both
taxonomic richness and evenness were maximized at an intermediate
intensity, around 20-25\% of AGB lost, and this along the 30 years
following disturbance.

Regarding community functional trajectories, no marked differences were
observed among the post-disturbance trajectories (\ref{fig:DivTaxo}).
Whenever the disturbance intensity, functional richness and evenness
first increased with the recruitment of pioneers that were functionally
highly different from the pre-disturbance community
\citep{Denslow1980, Molino2001}. Then, after 15 to 20 years community
functional richness and evenness started to decrease, with the decline
of the first established species. Right after disturbance, short-lived
pioneers benefited from the resources made available and prevented the
establishment of other species. These pioneers then declined, which
decreased the functional richness and evenness \citep{Walker2009}.

Considering different times after disturbance, both functional richness
and evenness kept increasing with the disturbance intensity and did not
display humped-shaped trajectories (\ref{fig:IDHplot}).
